Exploring the composition of lithic assemblages in Mesolithic south-eastern Norway

### 2021-05-10 analysis/paper/paper.Rmd
Switching to word document for now

---
title: "Exploring the composition of lithic assemblages in Mesolithic south-eastern Norway"
author:
  - Isak Roalkvam:
      email: isak.roalkvam@iakh.uio.no
      institute: [UiO]
      correspondence: true
institute:
  - UiO: University of Oslo
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
    bookdown::html_document2:
      fig_caption: yes
      reference_docx: "../templates/template.docx" # Insert path for the DOCX file
      pandoc_args:
      - --lua-filter=../templates/scholarly-metadata.lua
      - --lua-filter=../templates/author-info-blocks.lua
      - --lua-filter=../templates/pagebreak.lua
bibliography: references.bib
csl: "../templates/journal-of-archaeological-science.csl" # Insert path for the bib-style
---

@@

```{r dist, fig.cap = "A) Spatial and B) temporal distribution of the sites chosen for analysis. The date ranges follow those given in the original excavation reports", fig.height = 10, fig.width = 12}
# Create a bounding box around site points to limit the overview map
sitbbox <- st_bbox(site_pts)
sitbbox[1:2] <- sitbbox[1:2] - 600000
sitbbox[3:4] <- sitbbox[3:4] + 600000
boundingpoly <- st_as_sf(st_as_sfc(sitbbox))

@@

yrs <- seq(11500, 5000, -500)
temp_dist <- ggplot(sites, aes(x = yrs, y = reorder(site, mean_date))) +
  geom_segment(aes(x = reported_start_bc, xend = reported_end_bc,
                   yend = reorder(site, mean_date),
                   colour = excavation_project), size = 1.7) +
  xlab("Cal. BC") +
  scale_x_reverse() +
  scale_colour_manual(labels = c("E18 Bommestad-Sky", "E18 Brunlanes",
                               "E18 Rugtvedt-Dørdal", "Vestfoldbaneprosjektet"),
                     values = c("springgreen4", "royalblue2",
                                "gold1", "firebrick3")) +
  labs(title = "B", colour = "Excavation project") +
  theme_classic() +
  theme(axis.title.y=element_blank(),
        axis.text.y=element_blank(),
        axis.ticks.y=element_blank(),
        legend.position = "bottom") +
  guides(colour =  guide_legend(title.position = "top", title.hjust = 0.5))

@@

@@

The date ranges provided for the sites are taken from the original excavation reports, and are the result of a consideration of typological indicators, dating with reference to shoreline displacement, and radiocarbon dates where these are available. Employing these date-ranges was deemed sufficient, as the main purpose is to get at general tendencies in the material, and not fine-grained chronological developments.

@@

While the definition of curation, its archaeological correlates, and behavioural implications have been widely discussed [e.g. @bamforth1986, @shott1996], following not least from the ambiguous definition first put forward by @binford1979, some concrete operationalisation of the terms have been suggested. Through a series of studies Barton and collegues [e.g. -@barton1999; @barton2014; @clark2017] have shown that the relationship between lithic volumetric density and relative frequency of retouched artefacts provide consistent results across a wide range of chronological and cultural context. This measure is taken to reflect degree of curation associated with the assemblages, which in turn is mainly to follow from variation in mobility patterns associated with the sites [@barton2014]. Higher degree of mobility would mean a higher dependency on the artefacts and the material people could bring with them, and dimensions such as weight, reliabilty, repairability, and the degree to which artefacts could be manipulated to fulfill a wide range of tasks are therefore assumed to have been factors of concern. From this it follows that the empirical expectation for short-term camps is a curated technological organisation with higher relative frequency of retouched artefacts, and a lower overall density of lithics [@clark2017]. More time spent in a single location, on the other hand, is in this model assumed to lead to better control of raw-material availability and to allow for its accumulation. This should in turn lead to an expedient technological organisation with reduced necessity for the conservation of lithics and extensive use of retouch. The empirical expectation for lower degree of mobility is therefore relatively high density of lithics and a low relative frequency of retouched artefacts. A further empirical expectation of an expendient tehcnological organisation is a higher number of cores and unretouched flakes and blades [@clark2017].

@@

lithics2 <- sites %>% 
  select(-n, -beach_flint_flint_nodule, -hammerstone, -excavation_project,
         -reported_start_bc, -reported_end_bc, -askeladden_id) %>% 
  # The reason for not using .keep below is for the option to 
  #retain artefact_count
  mutate(chip_freq =  sum(chip, chip_other) / artefact_count,
         core_freq = sum(core, core_other) / artefact_count,
         blanks_freq = sum(blade, microblade, fragment, flake, flake_other, 
                     fragment_other, blade_other,  microblade_other) / artefact_count,
         secondary_freq =  sum(axe_debitage, axe, chisel, 
                microlith, tanged_point, transverse_point, single_edged_point,
                hognipen_point, projectile_undefined, microburin, scraper, 
                knife, burin, drill, blade_retouch, microblade_retouch, 
                chip_retouch, fragment_retouch, flake_retouch, axe_other,
                axe_debitage_other, chisel_other, club_hatchet_other, 
                grinding_slab_stone, projectile_point_other, knife_other,
                scraper_other, drill_other, microblade_retouch_other, 
                blade_retouch_other, fragment_retouch_other,
                flake_retouch_other) / artefact_count,
         nonflint_freq = nonflint_count / artefact_count,
         lithic_dens = artefact_count / excavated_m3) %>%
  select(-blanks_lithics, -secondary_lithics, -excavated_m3, -artefact_count,
         -flint_count, -nonflint_count, -core, -core_fragment, -chip, 
         -chip_other, -core_other, -core_fragment_other, -dating_method) %>% 
  column_to_rownames("site")

@@

```{r cor-plot, fig.cap = "Correlation plot showing transformation of skewed variables for the PCA, and the linear correlation between these", fig.height = 9.5, fig.width = 10}
# Transform the skewed variables - sqrt() is used where there are zero values,
# and seems to provide an adequate result.
lithicstr <- lithics2 %>%
  # Transform skewed variables and reorder
  mutate("Mean date cal. BC*-1" = mean_date * -1,
          "log10(lithic_dens)" = log10(lithic_dens),
         "log10(secondary_freq)" = log10(secondary_freq),
         "core_freq^(1/3)" = core_freq^(1/3),
         "nonflint_freq^(1/3)" = nonflint_freq^(1/3),  .keep = "unused") %>% 
  relocate("Mean date cal. BC*-1",
    "log10(lithic_dens)", "blanks_freq", "log10(secondary_freq)",
    "chip_freq", "core_freq^(1/3)" 
  )

corr_plot(lithicstr)
```

```{r pca-analysis, fig.cap = "PCA using variables related. Note that details on the transformation of the variables has been left out of the plot for clarity.", fig.height = 7, fig.width = 10}
# Rename the transformed variables again to avoid clutter in the plot. 
# Specify in figure caption that the transformed variables are used 
lithicspca <- lithicstr %>% 
  mutate(mean_date =  `Mean date cal. BC*-1` * -1, .keep = "unused") %>% 
  rename(
   lithic_dens = "log10(lithic_dens)",
   secondary_freq = "log10(secondary_freq)",
   core_freq = "core_freq^(1/3)",
   nonflint_freq = "nonflint_freq^(1/3)"
  )

pcalithics <- PCA(select(lithicspca, -mean_date),
                  scale.unit = TRUE, graph = FALSE)
fviz_pca(pcalithics, geom.ind = "point", fill.ind = lithicspca$mean_date / 1000, 
          repel = TRUE, col.var = "black", pointshape = 21,  pointsize = 2.9,
         title = "") + 
  scale_fill_gradient2(name = "ka BP",  low="yellow", mid="orange", 
                        high="red", midpoint = centdate) +
  theme(legend.position = "bottom")

@@

To visualise this more clearly, a curation index based on the two variables was devised by perfoming a min-max normalisation on the two variables, scaling them to take on values between 0 and 1. The values for artefact density was then made negative to reflect its relationship with degree of curation. The mean was then found for each site on these two normalised values. Plotting the results indicates a higher degree of curation among the earliest sites, which then falls and settles without any further major fluctuations.

@@

```{r eigenvalues, fig.cap = "Table showing the eigenvalues of the components."}
# Create a table showing the eigenvalues of the components
round(as.tibble(get_eig(pcalithics)), 2) %>% 
  rename("Eigenvalue" = eigenvalue, "Variance (%)" = variance.percent, 
         "Cumulative variance (%)" = cumulative.variance.percent) %>% 
  rownames_to_column(var = "Component") %>% 
gt(rowname_col = "row", auto_align = FALSE)
```

@@

@@

```{r index}
# Create curation index
minmax <- lithics2 %>% 
  mutate(dens =  ((lithic_dens - min(lithic_dens)) / (max(lithic_dens) - min(lithic_dens))),
         secondary = ((secondary_freq - min(secondary_freq)) / (max(secondary_freq) - min(secondary_freq))))

@@

w3 <- ggplot(indexdata, aes(x = mean_date, y = index)) + 
  geom_point() +
  labs(title = "C",
       x = "Mean date cal. BC",
       y = "Curation index") +
  scale_x_reverse() + 
  geom_smooth(method = "loess", col = "red") +
  theme_classic() 
w3
```
# Discussion

The results of the CA does appear to align reasonably well with previous research that has been based on more informal analyses of artefact types, indicating that the employed artefact categories are capturing some behaviorally meaningful patterning that changes over time. However, as several authors have pointed out in relation to the application of CA for seriation purposes, 'time is not the only dimension'. The developments evident here do also most certainly correspond to more pervasive cultural change than a purely typo-chronological development of artefact morphology, highlighted perhaps most clearly by the influence of the later Mesolithic axe production sites. Unpicking and aligning the signficance of these chronological patterns with any specific behavioural dimensions using the CA results is, however, another task entirely. This follows most clearly from the fact that for the most part we lack even a most basic understanding of what any individual lithic object has been used for, leaving the behavioural significance of the employed units of analysis unclear (see section, above).

>>>>

---
title: "Exploring the composition of lithic assemblages in Mesolithic south-eastern Norway"
author:
  - Isak Roalkvam:
      email: isak.roalkvam@iakh.uio.no
      institute: [UiO]
      correspondence: true
institute:
  - UiO: University of Oslo
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
    bookdown::word_document2:
      fig_caption: yes
      reference_docx: "../templates/template.docx" # Insert path for the DOCX file
      pandoc_args:
      - --lua-filter=../templates/scholarly-metadata.lua
      - --lua-filter=../templates/author-info-blocks.lua
      - --lua-filter=../templates/pagebreak.lua
bibliography: references.bib
csl: "../templates/journal-of-archaeological-science.csl" # Insert path for the bib-style
---

@@

```{r dist, fig.cap = "A) Spatial and B) temporal distribution of the sites chosen for analysis. Radiocarbon probability density functions are given as the posterior sum. Solid lines indicate that the site has been dated with reference to relative sea-level change and typological indicators. These follow the original reports.", fig.height = 11, fig.fullwidth = TRUE}
# Create a bounding box around site points to limit the overview map
sitbbox <- st_bbox(site_pts)
sitbbox[1:2] <- sitbbox[1:2] - 600000
sitbbox[3:4] <- sitbbox[3:4] + 600000
boundingpoly <- st_as_sf(st_as_sfc(sitbbox))

@@

# Combine radiocarbon data and reported (effectively shoreline/typology) dates,
# removing the reported date if there are relevant 14C-dates.
datedata <- full_join(sites, rcarb_data, by = "site") %>% 
  mutate(reported_start_bc = ifelse(dating_method != "c14", reported_start_bc, NA),
         reported_end_bc = ifelse(dating_method != "c14", reported_end_bc, NA))

project_colors <- c("bommestad_sky" = "springgreen4",
                  "brunlanes" = "royalblue2", 
                  "rugtvedt_dordal" = "gold1",
                  "vestfoldbaneprosjektet" = "firebrick3")

# Assemble plot of date ranges
temp_dist <- ggplot(datedata, aes(x = mean_date, y = reorder(site, mean_date))) +
  geom_segment(aes(x = reported_start_bc, xend = reported_end_bc,
                   yend = site, colour = excavation_project.x,
                   fill = excavation_project.x), size = 1.8) +
  geom_ridgeline(aes(x = date_grid * -1, height = probabilities * 150,
                     fill = excavation_project.y, 
                     colour = excavation_project.x)) +
  scale_x_reverse() +
  scale_fill_manual(values = project_colors,
                    labels = c("E18 Bommestad-Sky","E18 Brunlanes",
                            "E18 Rugtvedt-Dørdal", "Vestfoldbaneprosjektet")) +
  scale_colour_manual(values = project_colors) +
  labs(title = "B", fill = "Excavation project") +
  xlab("Cal. BCE") +
  theme_classic() +
  theme(axis.title.y=element_blank(),
        axis.text.y=element_blank(),
        axis.ticks.y=element_blank(),
        legend.position = "bottom") +
  guides(fill =  guide_legend(title.position = "top", title.hjust = 0.5),
         colour = FALSE)

@@

#plot.margin = unit(c(0,135,0,135)
# top right, borrom , l

@@

@@

While the theoretical definition of curation, its archaeological correlates, and behavioural implications have been widely discussed [e.g. @bamforth1986; @shott1996; @surovell2009,9--13], following not least from the ambiguous definition first put forward by @binford1979. Some concrete operationalisation of the terms have, however, been suggested. Through a series of studies Barton and collegues [e.g. -@barton1999; @barton2014; @clark2017] have shown that the relationship between lithic volumetric density and relative frequency of retouched artefacts provide consistent results across a wide range of chronological and cultural context, and is measure taken to reflect degree of curation associated with the assemblages. which in turn is mainly to follow from variation in mobility patterns associated with the sites [@barton2014]. Higher degree of mobility would mean a higher dependency on the artefacts and the material people could bring with them, and dimensions such as weight, reliabilty, repairability, and the degree to which artefacts could be manipulated to fulfill a wide range of tasks are therefore assumed to have been factors of concern. From this it follows that the empirical expectation for short-term camps is a curated technological organisation with higher relative frequency of retouched artefacts, and a lower overall density of lithics [@clark2017]. More time spent in a single location, on the other hand, is in this model assumed to lead to better control of raw-material availability and to allow for its accumulation. This should in turn lead to an expedient technological organisation with reduced necessity for the conservation of lithics and extensive use of retouch. The empirical expectation for lower degree of mobility is therefore relatively high density of lithics and a low relative frequency of retouched artefacts. A further empirical expectation of an expendient tehcnological organisation is as a consequence also a higher number of cores and unretouched flakes and blades [@clark2017].

@@

lithics2 <- sites %>% 
  select(-n, -beach_flint_flint_nodule, -hammerstone, -excavation_project,
          -askeladden_id) %>% 
  # The reason for not using .keep below is for the option to 
  #retain artefact_count
  mutate(chip_freq =  sum(chip, chip_other) / artefact_count,
         core_freq = sum(core, core_other) / artefact_count,
         blanks_freq = sum(blade, microblade, fragment, flake, flake_other, 
                     fragment_other, blade_other,  microblade_other) / artefact_count,
         secondary_freq =  sum(axe_debitage, axe, chisel, 
                microlith, tanged_point, transverse_point, single_edged_point,
                hognipen_point, projectile_undefined, microburin, scraper, 
                knife, burin, drill, blade_retouch, microblade_retouch, 
                chip_retouch, fragment_retouch, flake_retouch, axe_other,
                axe_debitage_other, chisel_other, club_hatchet_other, 
                grinding_slab_stone, projectile_point_other, knife_other,
                scraper_other, drill_other, microblade_retouch_other, 
                blade_retouch_other, fragment_retouch_other,
                flake_retouch_other) / artefact_count,
         nonflint_freq = nonflint_count / artefact_count,
         lithic_dens = artefact_count / excavated_m3) %>%
  select(-blanks_lithics, -secondary_lithics, -excavated_m3, -artefact_count,
         -flint_count, -nonflint_count, -core, -core_fragment, -chip, 
         -chip_other, -core_other, -core_fragment_other) %>% 
  column_to_rownames("site")

@@

```{r cor-plot, fig.cap = "Correlation plot showing transformation of skewed variables for the PCA, and the linear correlation between these", fig.height = 9.5, fig.width = 10}
# Transform the skewed variables - sqrt() is used where there are zero values,
# and seems to provide an adequate result.
lithicstr <- lithics2 %>%
  # Transform skewed variables and reorder
  mutate("Mean date cal. BCE*-1" = mean_date * -1,
          "log10(lithic_dens)" = log10(lithic_dens),
         "log10(secondary_freq)" = log10(secondary_freq),
         "core_freq^(1/3)" = core_freq^(1/3),
         "nonflint_freq^(1/3)" = nonflint_freq^(1/3),  .keep = "unused") %>% 
  relocate("Mean date cal. BCE*-1",
    "log10(lithic_dens)", "blanks_freq", "log10(secondary_freq)",
    "chip_freq", "core_freq^(1/3)" 
  )

corr_plot(select(lithicstr, -reported_start_bc, -reported_end_bc, -dating_method))
```

```{r pca-analysis, fig.cap = "PCA using variables related. Note that details on the transformation of the variables has been left out of the plot for clarity.", fig.height = 7, fig.width = 10}
# Rename the transformed variables again to avoid clutter in the plot. 
# Specify in figure caption that the transformed variables are used 
lithicspca <- lithicstr %>% 
  mutate(mean_date =  `Mean date cal. BCE*-1` * -1, .keep = "unused") %>% 
  rename(
   lithic_dens = "log10(lithic_dens)",
   secondary_freq = "log10(secondary_freq)",
   core_freq = "core_freq^(1/3)",
   nonflint_freq = "nonflint_freq^(1/3)"
  )

pcalithics <- PCA(select(lithicspca, -reported_start_bc, -dating_method,
                         -reported_end_bc, -mean_date),
                  scale.unit = TRUE, graph = FALSE)
fviz_pca(pcalithics, geom.ind = "point", fill.ind = lithicspca$mean_date / 1000, 
          repel = TRUE, col.var = "black", pointshape = 21,  pointsize = 2.9,
         title = "") + 
  scale_fill_gradient2(name = "ka BP",  low="yellow", mid="orange", 
                        high="red", midpoint = centdate) +
  theme(legend.position = "bottom")

@@

@@

```{r eigenvalues, fig.cap = "Table showing the eigenvalues of the components."}
# Create a table showing the eigenvalues of the components
round(as_tibble(get_eig(pcalithics)), 2) %>% 
  rename("Eigenvalue" = eigenvalue, "Variance (%)" = variance.percent, 
         "Cumulative variance (%)" = cumulative.variance.percent) %>% 
  rownames_to_column(var = "Component") %>% 
gt(rowname_col = "row", auto_align = FALSE)
```

@@

To visualise this temporal trend more clearly, a curation index based on the two variables was devised by first performing a min-max normalisation of the two variables, scaling them to take on values between 0 and 1. The values for artefact density was then made negative to reflect its relationship with degree of curation. The mean was then found for each site on these two normalised values. To account for the temporal uncertainty associated with the sites, a simulation based approach was also adopted [e.g. @crema2012; @baxter2016; @orton2017]. A LOESS curve was fit to the data for each simulation, where the age of each site was drawn from the date ranges associated with the sites. For sites with radiocarbon age determinations the dates were drawn from the associated posterior probability density functions, while dates for sites dated with reference to relative sea-level change and typology were drawn uniformly from the associated date ranges. This simulation was repeated `r nsim` times, the results of which is visualed in . Disregarding the edge-effects at either end of the plot, the general tendency is a relatively high degree of curation among the earlier sites, followed by a marked drop around 8000 BCE before remaining relatively stable until the end of the Mesolithic. It is also interesting to note that the variation in degree of curation is higher after 8000 BCE, potentially reflecting variation in site types and associated mobility patterns.

@@

```{r index}
# Create curation index
minmax <- lithics2 %>% 
   rownames_to_column(var = "site") %>% 
  mutate(dens =  ((lithic_dens - min(lithic_dens)) / (max(lithic_dens) - min(lithic_dens))),
         secondary = ((secondary_freq - min(secondary_freq)) / (max(secondary_freq) - min(secondary_freq))))

@@

# Curation index plot simply using mean site age (not used)
w3 <- ggplot(indexdata, aes(x = mean_date, y = index)) + 
  geom_point() +
  labs(title = "C",
       x = "Mean date cal. BCE",
       y = "Curation index") +
  scale_x_reverse() + 
  geom_smooth(method = "loess", col = "red") +
  theme_classic() 
#w3
```

```{r simage, cache = TRUE}
agedata <- full_join(indexdata, rcarb_data, by = "site")

# Define function to draw sample from either 14C probability distribution,
# or uniform typological/displacement-inferred date range.
sample_date <- function(site){

  # If the dating method is 14C, draw a sample using the associated posterior
  # pdf
  if(site["dating_method"][1,] == "c14"){
    samplingframe <- list(x = site[["date_grid"]], y = site[["probabilities"]])
    sdate <- with(samplingframe, sample(x, size = 1, prob = y))
  # If not, draw a random year from the reported date range (uniform probability)
  } else {
    sdate <- sample(as.integer(site["reported_start_bc"]*-1):as.integer(site["reported_end_bc"]*-1), 1)
  }
  # Return sampled date with site name as rowname. 
  sdate <- data.frame(sdate)
  row.names(sdate) <- site[["site"]][1]
  return(sdate)
}

# Specify number of simulations
nsim <- 1000

# Create data frame to hold simulation results.
results <- data.frame(matrix(NA, ncol = nsim, nrow = nrow(sites))) 
# Insert the curation index value for each site
results[,"index"] <- agedata %>% 
  group_by(site) %>% 
  slice(1) %>% 
  ungroup() %>% 
  select(index) %>% 
  c()

# Simulation loop
for (i in 1:nsim){
  agesamples <- agedata %>% 
    group_by(site) %>% 
    group_map(~ sample_date(.), .keep = TRUE) %>% 
    do.call(rbind, .)

  results[,i] <- agesamples
  row.names(results) <- row.names(agesamples)
}

# Apply pivot_longer to make the results amenable to plotting
resultslong <- results %>% 
  rownames_to_column(var = "site") %>% 
  pivot_longer(
    cols = starts_with("X"),
    names_to = "sim", 
    values_to = "age"
  )

# Call to plot, grouping loess fitting by simulation runs
ggplot(resultslong, aes(x = age, y = index)) + 
  geom_smooth(aes(group = sim), method = "loess", fill = "grey", col = "black", alpha = 0.05, se = TRUE) +
  geom_point(size = 0.5, alpha = 0.005) +
  xlab("Cal. BCE") +
  ylab("Curation index") +
  theme_classic()

# Discussion

The results of the CA does appear to align reasonably well with previous research that has been based on more informal analyses of artefact types, indicating that the employed artefact categories are capturing some behaviorally meaningful patterning that changes over time. However, as several authors have pointed out in relation to the application of CA for seriation purposes, 'time is not the only dimension'. The developments evident here do also most certainly correspond to more pervasive cultural change than a purely typo-chronological development of artefact morphology, highlighted perhaps most clearly by the influence of the later Mesolithic axe production sites. Unpicking and aligning the significance of these chronological patterns with any specific behavioural dimensions using the CA results is, however, another task entirely. This follows most clearly from the fact that for the most part we lack even a most basic understanding of what any individual lithic object has been used for, leaving the behavioural significance of the employed units of analysis unclear (see section, above).
